%!TEX root = ../main.tex
\section*{References}
\addcontentsline{toc}{section}{References}

\begin{enumerate}[label={[\arabic*]}, leftmargin=1.2cm]
\item \label{bib:bruceslam} J.~Wang, F.~Chen, Y.~Huang, J.~McConnell, T.~Shan, B.~Englot,
  \emph{Virtual Maps for Autonomous Exploration of Cluttered Underwater Environments},
  IEEE Journal of Oceanic Engineering, 2022 (the underlying SLAM is the
  ``Bruce-SLAM'' used here). arXiv:2202.08359.
\item \label{bib:sonarcontext} H.~Kim, G.~Kang, S.~Jeong, S.~Ma, Y.~Cho,
  \emph{Robust Imaging Sonar-based Place Recognition and Localization in Underwater
  Environments}, ICRA 2023.
\item \label{bib:diso} S.~Xu, K.~Zhang, Z.~Hong, Y.~Liu, S.~Wang,
  \emph{DISO: Direct Imaging Sonar Odometry}, ICRA 2024.
\item \label{bib:isopot} J.~Samec et al.,
  \emph{ISOPoT: Imaging Sonar Odometry by Point Tracking}, arXiv:2606.23006, 2026.
\item \label{bib:insusbldvl} \emph{A Robust INS/USBL/DVL Integrated Navigation Algorithm Using Graph
  Optimization}, Sensors 23(2):916, 2023.
\item \label{bib:pcm} J.~Mangelson, D.~Dominic, R.~Eustice, R.~Vasudevan,
  \emph{Pairwise Consistent Measurement Set Maximization for Robust Multi-robot Map
  Merging}, ICRA 2018.
\item \label{bib:umeyama} S.~Umeyama,
  \emph{Least-Squares Estimation of Transformation Parameters Between Two Point
  Patterns}, IEEE TPAMI, 1991.
\item \label{bib:aracati} M.~dos Santos et al., ARACATI 2017 dataset —
  \url{https://github.com/matheusbg8/aracati2017}.
\item \label{bib:isam2} M.~Kaess et al.,
  \emph{iSAM2: Incremental Smoothing and Mapping Using the Bayes Tree}, IJRR 2012.
\item \label{bib:sonic} S.~Gode, A.~Hinduja, M.~Kaess,
  \emph{SONIC: Sonar Image Correspondence using Pose Supervised Learning for Imaging
  Sonars}, arXiv:2310.15023, 2024.
\item \label{bib:holoocean} E.~Potokar, S.~Ashford, M.~Kaess, J.~Mangelson,
  \emph{HoloOcean: An Underwater Robotics Simulator}, ICRA 2022.
\item \label{bib:mcconnell} J.~McConnell, J.~D.~Martin, B.~Englot,
  \emph{Fusing Concurrent Orthogonal Wide-aperture Sonar Images for Dense Underwater
  3D Reconstruction}, IROS 2020.
\end{enumerate}
